---
title: "Ghanadan_Lab2"
author: "Linus Ghanadan"
date: "1-18-2024"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
# Load packages
library(knitr)

# Load variables from lab 1
source(knitr::purl("~/Documents/MEDS/eds232-ml-code/eds232-ml-labs/Lab1.Rmd"))
pumpkins_train <- readRDS("~/Documents/MEDS/eds232-ml-code/eds232-ml-labs/pumpkins_train.rds")
```

Today we will be continuing the pumpkin case study from last week. We will be using the data that you cleaned and split last time (pumpkins_train) and will be comparing our results today to those you have already obtained. Open and run your Lab 1.Rmd as a first step so those objects are available in your Environment.

Once you have done that, we'll start today's lab by specifying a recipe for a polynomial model.  First we specify a recipe that identifies our variables and data, converts the package variable to a numerical form, and then adds a polynomial effect with step_poly()

```{r}
# Specify a recipe
poly_pumpkins_recipe <-
  recipe(price ~ package, data = pumpkins_train) %>%
  step_integer(all_predictors(), zero_based = TRUE) %>% 
  step_poly(all_predictors(), degree = 4)
```

How did that work? Later we will learn about model tuning that will let us do things like find the optimal value for degree.  For now, we'd like to have a flexible model, so we'll use a relatively large value.

Polynomial regression is still linear regression, so our model specification looks similar to before.

```{r}
# Create a model specification called poly_spec
poly_spec <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")
```
Question 1: Now take the recipe and model specification that just created and bundle them into a workflow called poly_wf.

```{r}
# Bundle recipe and model spec into a workflow
poly_wf <- workflow() %>% 
  add_recipe(poly_pumpkins_recipe) %>% 
  add_model(poly_spec)
```

Question 2: fit a model to the pumpkins_train data using your workflow and assign it to poly_wf_fit
```{r}
# Create a model
poly_wf_fit <- poly_wf %>% 
  fit(data = pumpkins_train)
```

```{r}
# Print learned model coefficients
poly_wf_fit
```


```{r}
# Make price predictions on test data
poly_results <- poly_wf_fit %>% predict(new_data = pumpkins_test) %>% 
  bind_cols(pumpkins_test %>% select(c(package, price))) %>% 
  relocate(.pred, .after = last_col())

# Print the results
poly_results %>% 
  slice_head(n = 10)
```

Now let's evaluate how the model performed on the test_set using yardstick::metrics().
```{r}
metrics(data = poly_results, truth = price, estimate = .pred)
```
[Question 3: How do the performance metrics differ between the linear model from last week and the polynomial model we fit today?  Which model performs better on predicting the price of different packages of pumpkins?
[The Root Mean Square Error (RMSE) of our polynomial model (3.27) is smaller than that from our linear model (7.23). This means that for our polynomial model, our prediction is, on average, 3.27 dollars off, compared to 7.23 dollars off for the linear model. In addition, the Coefficient of Determination of our polynomial model (0.89) is larger than that from our linear model (0.49). This means that 89% of the variation in price can be explained by our polynomial model, while only 49% of this variation can be explained by our linear model. Thus, because our polynomial model has a lower RMSE and higher R-squared, it performs better on predicting the price of different packages of pumpkins.]{style="color:navy;"}

Let's visualize our model results.  First prep the results by binding the encoded package variable to them.
```{r}
# Bind encoded package column to the results
poly_results <- poly_results %>% 
  bind_cols(package_encode %>% 
              rename(package_integer = package)) %>% 
  relocate(package_integer, .after = package)


# Print new results data frame
poly_results %>% 
  slice_head(n = 5)
```

OK, now let's take a look! 

Question 4: Create a scatter plot that takes the poly_results and plots package vs. price.  Then draw a line showing our model's predicted values (.pred). Hint: you'll need separate geoms for the data points and the prediction line.
```{r}
# Make a scatter plot
poly_results %>% 
  ggplot(mapping = aes(x = package_integer, y = price)) +
   geom_point(size = 1.6) +
   # Overlay a line of best fit
   geom_line(aes(y = .pred), color = "orange", linewidth = 1.2) +
   xlab("package")

```

You can see that a curved line fits your data much better.

Question 5: Now make a smoother line by using geom_smooth instead of geom_line and passing it a polynomial formula like this:
geom_smooth(method = lm, formula = y ~ poly(x, degree = 3), color = "midnightblue", size = 1.2, se = FALSE)

```{r}
# Make a smoother scatter plot 
poly_results %>% 
  ggplot(mapping = aes(x = package_integer, y = price)) +
   geom_point(size = 1.6) +
   # Overlay a line of best fit
   geom_smooth(method = lm, formula = y ~ poly(x, degree = 3), color = "midnightblue", size = 1.2, se = FALSE) +
   xlab("package")
```

OK, now it's your turn to go through the process one more time.
 
Additional assignment components :
6. Choose a new predictor variable (anything not involving package type) in this dataset.
[Variety]{style="color:navy;"}

7. Determine its correlation with the outcome variable (price).  (Remember we calculated a correlation matrix last week)
```{r}
# Correlation between price and variety
cor(baked_pumpkins$variety, baked_pumpkins$price)
```

8. Create and test a model for your new predictor:
  - Create a recipe
  - Build a model specification (linear or polynomial)
  - Bundle the recipe and model specification into a workflow
  - Create a model by fitting the workflow
  - Evaluate model performance on the test data
  - Create a visualization of model performance
  
```{r}
# Specify a recipe
new_lm_pumpkins_recipe <-
  recipe(price ~ variety, data = pumpkins_train) %>%
  step_integer(all_predictors(), zero_based = TRUE)

# Create a linear model specification
new_lm_spec <- linear_reg() %>% 
  set_engine("lm") %>% 
  set_mode("regression")

# Hold modeling components in a workflow
new_lm_wf <- workflow() %>% 
  add_recipe(new_lm_pumpkins_recipe) %>% 
  add_model(lm_spec)

new_lm_wf
```

```{r}
# Train the model
new_lm_wf_fit <- new_lm_wf %>% 
  fit(data = pumpkins_train)

new_lm_wf_fit
```

```{r}
# Make predictions for the test set
predictions <- new_lm_wf_fit %>% 
  predict(new_data = pumpkins_test)


# Bind predictions to the test set
new_lm_results <- pumpkins_test %>% 
  select(c(variety, price)) %>% 
  bind_cols(predictions)

# Print the first ten rows of the tibble
new_lm_results %>% 
  slice_head(n = 10)
```

```{r}
# Evaluate performance of linear regression
metrics(data = new_lm_results,
        truth = price,
        estimate = .pred)
```

```{r}
# Make a scatter plot
new_lm_results %>%
  ggplot(mapping = aes(x = variety, y = price)) +
   geom_point(size = 1.6) +
   # Overlay a line of best fit
   geom_line(aes(y = .pred), color = "orange", linewidth = 1.2) +
   xlab("variety")
```

  
Lab 2 due 1/24 at 11:59 PM
