---
title: "Ghanadan_Lab4"
author: "Linus Ghanadan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(skimr)
library(tidymodels)
library(caret)
library(corrplot)
library(boot)
```

## Lab 4: Fire and Tree Mortality

The database we'll be working with today includes 36066 observations of individual trees involved in prescribed fires and wildfires occurring over 35 years, from 1981 to 2016. It is a subset of a larger fire and tree mortality database from the US Forest Service (see data description for the full database here: [link](https://www.nature.com/articles/s41597-020-0522-7#Sec10)). Our goal today is to predict the likelihood of tree mortality after a fire.

### Data Exploration

Outcome variable: *yr1status* = tree status (0=alive, 1=dead) assessed one year post-fire.

Predictors: *YrFireName, Species, Genus_species, DBH_cm, CVS_percent, BCHM_m, BTL* (Information on these variables available in the database metadata ([link](https://www.fs.usda.gov/rds/archive/products/RDS-2020-0001-2/_metadata_RDS-2020-0001-2.html))).

```{r}
trees_dat <- read_csv(file = "https://raw.githubusercontent.com/MaRo406/eds-232-machine-learning/main/data/trees-dat.csv")
```

> Question 1: Recode all the predictors to a zero_based integer form

```{r}
# specify a recipe
trees_recipe <- trees_dat %>% 
  recipe(yr1status ~ .) %>%
  step_integer(all_string_predictors(), zero_based = TRUE) %>% 
  prep()
```


### Data Splitting

> Question 2: Create trees_training (70%) and trees_test (30%) splits for the modeling

```{r}
# create training (70%) and test (30%) sets for the trees data
set.seed(123)  # for reproducibility 
trees_split <- initial_split(trees_dat, prop = .70)
trees_train <- training(trees_split)
trees_test  <- testing(trees_split)
```

> Question 3: How many observations are we using for training with this split?

```{r}
nrow(trees_train)
```

[There are 25246 observations in our training data.]{style="color:navy;"}

### Simple Logistic Regression 

Let's start our modeling effort with some simple models: one predictor and one outcome each.

> Question 4: Choose the three predictors that most highly correlate with our outcome variable for further investigation.

```{r}
# bake recipe with training data
trees_baked_train <- bake(trees_recipe, new_data = trees_train)
```


```{r}
# Obtain correlation matrix
corr_mat <- cor(trees_baked_train)

# Make a correlation plot between the variables
corrplot(corr_mat, method = "shade", shade.col = NA, tl.col = "black", tl.srt = 45, addCoef.col = "black", cl.pos = "n", order = "original")
```

[The three variables most correlated with *yr1status* are *CVS_percent* (0.68), *BCHM_m* (0.42), and *DBH_cm* (-0.3).]{style="color:navy;"}

> Question 5: Use glm() to fit three simple logistic regression models, one for each of the predictors you identified.

```{r}
# Build model for CVS_percent
model_cvs <- glm(data = trees_baked_train, yr1status ~ CVS_percent, family = "binomial")
  
# Build model for BCHM_m
model_bchm <- glm(data = trees_baked_train, yr1status ~ BCHM_m, family = "binomial")

# Build model for DBH_cm
model_dbh <- glm(data = trees_baked_train, yr1status ~ DBH_cm, family = "binomial")

# tidy models
tidy(model_cvs)
tidy(model_bchm)
tidy(model_dbh)

# exponentiate the coefficients from model objects for interpretation., giving changes in odds of tree being dead one year after fire
exp(coef(model_cvs))
exp(coef(model_bchm))
exp(coef(model_dbh))

```


### Interpret the Coefficients 

We aren't always interested in or able to interpret the model coefficients in a machine learning task. Often predictive accuracy is all we care about.

> Question 6: That said, take a stab at interpreting our model coefficients now.

[The odds of a tree being dead one year after a fire multiply 1.08-fold for every 1 additional percent of crown volume that was scorched.]{style="color:navy;"}

[The odds of a tree being dead one year after a fire multiply 1.235-fold for every 1 additional meter of vertical height with maximum bark char.]{style="color:navy;"}

[The odds of a tree being dead one year after a fire multiply 0.94-fold for every 1 additional centimeter in diameter at breast height.]{style="color:navy;"}

> Question 7: Now let's visualize the results from these models. Plot the fit to the training data of each model.

```{r}
# Plot CVS model with training data
ggplot(trees_baked_train, aes(x = CVS_percent, y = yr1status)) +
  geom_point() +
  stat_smooth(method = "glm", se = TRUE, method.arguments = list(family = "binomial"))
```

```{r}
# Plot BCHM model with training data
ggplot(trees_baked_train, aes(x = BCHM_m, y = yr1status)) +
  geom_point() +
  stat_smooth(method = "glm", se = TRUE, method.arguments = list(family = "binomial"))
```

```{r}
# Plot DBH model with training data
ggplot(trees_baked_train, aes(x = DBH_cm, y = yr1status)) +
  geom_point() +
  stat_smooth(method = "glm", se = TRUE, method.arguments = list(family = "binomial"))
```


### Multiple Logistic Regression

Let's not limit ourselves to a single-predictor model. More predictors might lead to better model performance.


> Question 8: Use glm() to fit a multiple logistic regression called "logistic_full", with all three of the predictors included. Which of these are significant in the resulting model?

```{r}
# build multiple logistic regression model with all three predictors
logistic_full <- glm(yr1status ~ CVS_percent + BCHM_m + DBH_cm, family = "binomial", data = trees_train)

# tidy model
tidy(logistic_full)
```


### Estimate Model Accuracy

Now we want to estimate our model's generalizability using resampling.

> Question 9: Use cross validation to assess model accuracy. Use caret::train() to fit four 10-fold cross-validated models (cv_model1, cv_model2, cv_model3, cv_model4) that correspond to each of the four models we've fit so far: three simple logistic regression models corresponding to each of the three key predictors (CVS_percent, DBH_cm, BCHM_m) and a multiple logistic regression model that combines all three predictors.

```{r}
# Define the control parameters for 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 10)

# Create a formula for each model
formula1 <- as.formula("yr1status ~ CVS_percent")
formula2 <- as.formula("yr1status ~ DBH_cm")
formula3 <- as.formula("yr1status ~ BCHM_m")
formula4 <- as.formula("yr1status ~ CVS_percent + DBH_cm + BCHM_m")

# Fit the cross-validated models
cv_model1 <- train(formula1, data = trees_baked_train, method = "glm", trControl = ctrl, family = "binomial")
cv_model2 <- train(formula2, data = trees_baked_train, method = "glm", trControl = ctrl, family = "binomial")
cv_model3 <- train(formula3, data = trees_baked_train, method = "glm", trControl = ctrl, family = "binomial")
cv_model4 <- train(formula4, data = trees_baked_train, method = "glm", trControl = ctrl, family = "binomial")

# Print the cross-validated results
print(cv_model1)
print(cv_model2)
print(cv_model3)
print(cv_model4)


```


> Question 10: Use caret::resamples() to extract then compare the classification accuracy for each model. (Hint: resamples() wont give you what you need unless you convert the outcome variable to factor form). Which model has the highest accuracy?

```{r}
# Convert the outcome variable to factor
trees_baked_train$yr1status <- as.factor(trees_baked_train$yr1status)

# Create a list of models
models_list <- list(cv_model1, cv_model2, cv_model3, cv_model4)

# Use resamples() to compare classification accuracy
resamp <- resamples(models_list)

# Print the summary of resampled results
summary(resamp)

```



Let's move forward with this single most accurate model.

> Question 11: Compute the confusion matrix and overall fraction of correct predictions by the model.

```{r}
# Make predictions on your data using the trained model
predictions <- predict(cv_model4, newdata = trees_baked_train)

# Create the confusion matrix
confusion_matrix <- confusionMatrix(predictions, reference = trees_baked_train$yr1status)

# Print the confusion matrix
print(confusion_matrix)

```


```{r}
# Compute the overall fraction of correct predictions
correct_predictions <- confusion_matrix$overall["Accuracy"]
print(paste("Overall Fraction of Correct Predictions:", correct_predictions))
```


> Question 12: Explain what the confusion matrix is telling you about the types of mistakes made by logistic regression.
>
> Question 13: What is the overall accuracy of the model? How is this calculated?

### Test Final Model

Alright, now we'll take our most accurate model and make predictions on some unseen data (the test data).

> Question 14: Now that we have identified our best model, evaluate it by running a prediction on the test data, trees_test.
>
> Question 15: How does the accuracy of this final model on the test data compare to its cross validation accuracy? Do you find this to be surprising? Why or why not?
